---
title: "QC Report — Metagenomics Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)

# ══════════════════════════════════════════════════════════════════════════════
# USER-ADJUSTABLE PARAMETERS
# ══════════════════════════════════════════════════════════════════════════════
# Samples with fewer reads than this after trimming will be flagged
MIN_READS <- 10000
```

```{r load_config, include=FALSE}
config_lines <- readLines("../config.sh")

extract_config <- function(lines, key) {
  line <- lines[grep(paste0("^", key, "="), lines)]
  if (length(line) == 0) return(NA)
  value <- gsub(paste0(key, "=|\""), "", line[1])
  value <- gsub("#.*$", "", value)
  trimws(value)
}

cfg_type          <- extract_config(config_lines, "TYPE")
cfg_remove_human  <- extract_config(config_lines, "REMOVE_HUMAN")
cfg_food_host_ref <- extract_config(config_lines, "FOOD_HOST_REF")
cfg_out_dir       <- extract_config(config_lines, "OUT_DIR")
```

```{r load_data, include=FALSE}
# ── Metadata ──────────────────────────────────────────────────────────────────
metadata_path <- "../metadata.tsv"
if (!file.exists(metadata_path)) {
  stop("metadata.tsv not found. Please add it to 04_aggregated/ before rendering.")
}
metadata <- read_tsv(metadata_path, col_types = cols(.default = "c")) %>%
  rename_with(tolower)

required_cols <- c("sequence_id", "sample_name", "sample_group")
missing_cols  <- setdiff(required_cols, colnames(metadata))
if (length(missing_cols) > 0) {
  stop(paste("metadata.tsv is missing columns:", paste(missing_cols, collapse = ", ")))
}

# ── Build colour palette from sample_group ────────────────────────────────────
groups        <- unique(metadata$sample_group)
non_blank     <- groups[groups != "blank"]
base_colours  <- c("#4E79A7", "#59A14F", "#F28E2B", "#B07AA1",
                   "#76B7B2", "#EDC948", "#FF9DA7", "#9C755F")
group_colours <- setNames(base_colours[seq_along(non_blank)], non_blank)
group_colours["blank"] <- "#AAAAAA"

# ── Hostile summary ───────────────────────────────────────────────────────────
hostile_path <- "hostile_summary.csv"
if (!file.exists(hostile_path)) {
  stop("hostile_summary.csv not found. Has 01b_qc_merge completed successfully?")
}
hostile <- read_csv(hostile_path, col_types = cols(.default = "c")) %>%
  mutate(
    reads_in                       = as.numeric(reads_in),
    reads_after_human              = as.numeric(reads_after_human),
    reads_out                      = as.numeric(reads_out),
    reads_removed_human            = as.numeric(reads_removed_human),
    reads_removed_proportion_human = as.numeric(reads_removed_proportion_human),
    reads_removed_food             = as.numeric(reads_removed_food),
    reads_removed_proportion_food  = as.numeric(reads_removed_proportion_food),
    food_host                      = as.character(food_host)
  )

# Derive food host summary variables used by section 2.2
any_food_removed <- any(hostile$food_host != "none" & !is.na(hostile$food_host))
unique_hosts     <- hostile %>%
  filter(food_host != "none", !is.na(food_host)) %>%
  pull(food_host) %>%
  unique() %>%
  sort()
unique_hosts_label <- paste(unique_hosts, collapse = ", ")

# ── Join metadata and hostile ─────────────────────────────────────────────────
data <- hostile %>%
  left_join(metadata, by = c("sample_name" = "sequence_id"))

unmatched <- data %>% filter(is.na(sample_group))
if (nrow(unmatched) > 0) {
  warning(paste(
    "The following samples are in hostile_summary.csv but not in metadata.tsv:",
    paste(unmatched$sample_name, collapse = ", ")
  ))
}

# Summary counts
n_total   <- nrow(metadata)
n_samples <- sum(metadata$sample_group != "blank", na.rm = TRUE)
n_blanks  <- sum(metadata$sample_group == "blank", na.rm = TRUE)

# ── MultiQC general stats ─────────────────────────────────────────────────────
multiqc_path <- "multiqc/multiqc_report_data/multiqc_general_stats.txt"
if (!file.exists(multiqc_path)) {
  stop("multiqc_general_stats.txt not found. Has 01b_qc_merge completed successfully?")
}
multiqc_raw <- read_tsv(multiqc_path, col_types = cols(.default = "c")) %>%
  rename(
    sample_raw      = Sample,
    pct_duplicates  = `FastQC_mqc-generalstats-fastqc-percent_duplicates`,
    pct_gc          = `FastQC_mqc-generalstats-fastqc-percent_gc`,
    avg_length      = `FastQC_mqc-generalstats-fastqc-avg_sequence_length`,
    median_length   = `FastQC_mqc-generalstats-fastqc-median_sequence_length`,
    pct_fails       = `FastQC_mqc-generalstats-fastqc-percent_fails`,
    total_sequences = `FastQC_mqc-generalstats-fastqc-total_sequences`
  ) %>%
  mutate(across(c(pct_duplicates, pct_gc, avg_length,
                  median_length, pct_fails, total_sequences), as.numeric))

# Keep R1 only and strip TrimGalore suffix to recover sequence_id
multiqc <- multiqc_raw %>%
  filter(grepl("_R1_", sample_raw)) %>%
  mutate(sequence_id = gsub("_hostremoved_R1_val_1$", "", sample_raw)) %>%
  left_join(metadata, by = "sequence_id")

# ── MultiQC FastQC per-module stats ───────────────────────────────────────────
fastqc_path <- "multiqc/multiqc_report_data/multiqc_fastqc.txt"
if (!file.exists(fastqc_path)) {
  stop("multiqc_fastqc.txt not found. Has 01b_qc_merge completed successfully?")
}
fastqc_modules <- read_tsv(fastqc_path, col_types = cols(.default = "c")) %>%
  filter(grepl("_R1_", Sample)) %>%
  mutate(sequence_id = gsub("_hostremoved_R1_val_1$", "", Sample)) %>%
  left_join(metadata, by = "sequence_id") %>%
  mutate(display_name = ifelse(is.na(sample_name), sequence_id, sample_name))

module_cols <- c(
  "basic_statistics", "per_base_sequence_quality", "per_sequence_quality_scores",
  "per_base_sequence_content", "per_sequence_gc_content", "per_base_n_content",
  "sequence_length_distribution", "sequence_duplication_levels",
  "overrepresented_sequences", "adapter_content"
)
```

---

# 1. Project Overview

```{r project_summary_table}
tibble(
  Parameter = c(
    "Sequencing type",
    "Human host removal",
    "Food host reference",
    "Total samples",
    "Biological samples",
    "Blank samples",
    "Report generated"
  ),
  Value = c(
    toupper(cfg_type),
    cfg_remove_human,
    ifelse(any_food_removed, unique_hosts_label, "not applied"),
    n_total,
    n_samples,
    n_blanks,
    as.character(Sys.Date())
  )
) %>%
  kable(col.names = c("Parameter", "Value"), align = "ll") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left"
  )
```

## Sample List

```{r sample_list_table}
table_data <- data %>%
  select(
    sequence_id  = sample_name,
    display_name = sample_name.y,
    group        = sample_group,
    reads_in,
    reads_out
  ) %>%
  mutate(
    reads_in  = formatC(reads_in,  format = "d", big.mark = ","),
    reads_out = formatC(reads_out, format = "d", big.mark = ","),
    group     = factor(group)
  ) %>%
  arrange(group, sequence_id)

blank_rows <- which(table_data$group == "blank")

table_data %>%
  kable(
    col.names = c("Sequence ID", "Sample Name", "Group",
                  "Raw Reads", "Reads After Host Removal"),
    align = "lllrr"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) %>%
  row_spec(blank_rows, background = "#f0f0f0")
```

---

# 2. Host Removal

## 2.1 Human Host Removal

```{r human_removal, results='asis'}
if (cfg_remove_human != "true") {
  cat("Human host removal was not applied in this run (`REMOVE_HUMAN=false`).\n")
} else {
  cat("Proportion of reads removed by Hostile (human reference) per sample.
Samples where the removal proportion is an outlier (beyond 2 standard deviations
from the mean of biological samples) are flagged in red.\n")
}
```

```{r human_removal_plot, fig.width=10, fig.height=5, eval=(cfg_remove_human == "true")}
human_data <- data %>%
  mutate(
    pct_removed  = reads_removed_proportion_human * 100,
    display_name = ifelse(is.na(sample_name.y), sample_name, sample_name.y),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group)
  )

# Outlier detection on biological samples only
bio_mean <- mean(human_data$pct_removed[human_data$sample_group != "blank"],
                 na.rm = TRUE)
bio_sd   <- sd(human_data$pct_removed[human_data$sample_group != "blank"],
               na.rm = TRUE)

human_data <- human_data %>%
  mutate(outlier = abs(pct_removed - bio_mean) > 2 * bio_sd &
                   sample_group != "blank")

plot_colours <- c(group_colours, "outlier" = "#E15759")

ggplot(human_data, aes(
    x    = reorder(display_name, pct_removed),
    y    = pct_removed,
    fill = ifelse(outlier, "outlier", sample_group)
  )) +
  geom_col() +
  geom_hline(yintercept = bio_mean, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = plot_colours, name = NULL) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "% Reads Removed",
    title = "Human host reads removed (human-t2t-hla reference)"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r human_removal_flags, results='asis', eval=(cfg_remove_human == "true")}
flags <- human_data %>%
  filter(outlier) %>%
  select(display_name, pct_removed) %>%
  mutate(pct_removed = round(pct_removed, 2))

if (nrow(flags) > 0) {
  cat("**⚠ Flagged samples (outlier human removal rate):**\n\n")
  flags %>%
    kable(col.names = c("Sample", "% Human Reads Removed"), align = "lr") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    print()
} else {
  cat("✓ No outlier samples detected for human read removal rate.\n")
}
```

## 2.2 Food Host Removal

```{r food_removal, results='asis'}
if (!any_food_removed) {
  cat("Food host removal was not applied in this run.\n")
} else {
  cat(paste0(
    "Proportion of reads removed by Hostile (food host reference) per sample, ",
    "after human removal. Food hosts removed: **", unique_hosts_label, "**. ",
    "Samples where the removal proportion is an outlier (beyond 2 standard deviations ",
    "from the mean within each food host group) are flagged in red.\n"
  ))
}
```

```{r food_removal_plot, fig.width=10, fig.height=5, eval=any_food_removed}
food_data <- data %>%
  mutate(
    pct_removed  = reads_removed_proportion_food * 100,
    display_name = ifelse(is.na(sample_name.y), sample_name, sample_name.y),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group),
    food_host    = ifelse(is.na(food_host) | food_host == "none",
                          "not applied", food_host)
  )

# Outlier detection per food host group, biological samples only
food_data <- food_data %>%
  group_by(food_host) %>%
  mutate(
    grp_mean = mean(pct_removed[sample_group != "blank"], na.rm = TRUE),
    grp_sd   = sd(pct_removed[sample_group != "blank"],   na.rm = TRUE),
    outlier  = abs(pct_removed - grp_mean) > 2 * grp_sd &
               sample_group != "blank" &
               food_host != "not applied"
  ) %>%
  ungroup()

plot_colours <- c(group_colours, "outlier" = "#E15759")

ggplot(food_data, aes(
    x    = reorder(display_name, pct_removed),
    y    = pct_removed,
    fill = ifelse(outlier, "outlier", sample_group)
  )) +
  geom_col() +
  scale_fill_manual(values = plot_colours, name = NULL) +
  facet_wrap(~ food_host, scales = "free_y", ncol = 1) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "% Reads Removed",
    title = paste0("Food host reads removed (", unique_hosts_label, ")")
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "#f0f0f0")
  )
```

```{r food_removal_flags, results='asis', eval=any_food_removed}
flags_food <- food_data %>%
  filter(outlier) %>%
  select(display_name, food_host, pct_removed) %>%
  mutate(pct_removed = round(pct_removed, 2)) %>%
  arrange(food_host, desc(pct_removed))

if (nrow(flags_food) > 0) {
  cat("**⚠ Flagged samples (outlier food host removal rate):**\n\n")
  flags_food %>%
    kable(col.names = c("Sample", "Food Host", "% Reads Removed"), align = "llr") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    print()
} else {
  cat(paste0("✓ No outlier samples detected across food host groups (",
             unique_hosts_label, ").\n"))
}
```

---

# 3. Read Quality After Trimming

## 3.1 Read Counts After Trimming

Total reads retained per sample after all host removal steps and quality trimming
(R1 only). Samples below the minimum read threshold
(`r format(MIN_READS, big.mark=",")` reads) are flagged in red.

```{r read_counts_plot, fig.width=10, fig.height=5}
reads_data <- multiqc %>%
  mutate(
    display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group),
    low_reads    = total_sequences < MIN_READS
  )

plot_colours_reads <- c(group_colours, "low_reads" = "#E15759")

ggplot(reads_data, aes(
    x    = reorder(display_name, total_sequences),
    y    = total_sequences / 1e6,
    fill = ifelse(low_reads, "low_reads", sample_group)
  )) +
  geom_col() +
  geom_hline(yintercept = MIN_READS / 1e6, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = plot_colours_reads, name = NULL) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "Total Reads (millions)",
    title = "Reads retained after host removal and trimming (R1)"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

## 3.2 GC Content and Duplication Rate

```{r gc_dup_plot, fig.width=12, fig.height=5}
gc_dup_data <- multiqc %>%
  mutate(
    display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group)
  )

p1 <- ggplot(gc_dup_data, aes(
    x    = reorder(display_name, pct_gc),
    y    = pct_gc,
    fill = sample_group
  )) +
  geom_col() +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = group_colours, name = NULL) +
  coord_flip() +
  labs(x = NULL, y = "% GC", title = "GC Content") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none")

p2 <- ggplot(gc_dup_data, aes(
    x    = reorder(display_name, pct_duplicates),
    y    = pct_duplicates,
    fill = sample_group
  )) +
  geom_col() +
  scale_fill_manual(values = group_colours, name = NULL) +
  coord_flip() +
  labs(x = NULL, y = "% Duplicates", title = "Duplication Rate") +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")

p1 + p2
```

---

# 4. Quality Flags

## 4.1 Samples Below Minimum Read Threshold

```{r low_reads_flag, results='asis'}
low_reads <- reads_data %>%
  filter(low_reads) %>%
  select(display_name, sample_group, total_sequences) %>%
  mutate(total_sequences = formatC(total_sequences, format = "d", big.mark = ","))

if (nrow(low_reads) > 0) {
  cat(paste0("**⚠ ", nrow(low_reads), " sample(s) below minimum read threshold (",
             format(MIN_READS, big.mark = ","), " reads):**\n\n"))
  low_reads %>%
    kable(col.names = c("Sample", "Group", "Total Reads"), align = "llr") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    row_spec(seq_len(nrow(low_reads)), background = "#f8d7da") %>%
    print()
} else {
  cat(paste0("✓ All samples are above the minimum read threshold (",
             format(MIN_READS, big.mark = ","), " reads).\n"))
}
```

## 4.2 FastQC Module Status

Pass/warn/fail status per FastQC module per sample (R1 only).
Failing modules are highlighted in red, warnings in orange.

```{r fastqc_module_heatmap, fig.width=12, fig.height=6}
heatmap_data <- fastqc_modules %>%
  select(display_name, all_of(module_cols)) %>%
  pivot_longer(
    cols      = all_of(module_cols),
    names_to  = "module",
    values_to = "status"
  ) %>%
  mutate(
    status = factor(status, levels = c("pass", "warn", "fail")),
    module = gsub("_", " ", module)
  )

status_colours <- c(
  "pass" = "#59A14F",
  "warn" = "#F28E2B",
  "fail" = "#E15759"
)

ggplot(heatmap_data, aes(x = module, y = display_name, fill = status)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_manual(values = status_colours, name = "Status", drop = FALSE) +
  labs(x = NULL, y = NULL, title = "FastQC Module Status (R1)") +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )
```

```{r fastqc_fail_summary, results='asis'}
fail_summary <- heatmap_data %>%
  filter(status == "fail") %>%
  group_by(module) %>%
  summarise(n_failing = n(), samples = paste(display_name, collapse = ", "),
            .groups = "drop") %>%
  arrange(desc(n_failing))

if (nrow(fail_summary) > 0) {
  cat("**⚠ Modules with failing samples:**\n\n")
  fail_summary %>%
    kable(col.names = c("Module", "# Failing Samples", "Samples"),
          align = "lrl") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE) %>%
    row_spec(seq_len(nrow(fail_summary)), background = "#f8d7da") %>%
    print()
} else {
  cat("✓ No samples have failing FastQC modules.\n")
}
```

## 4.3 Blank Sample Comparison

Read counts and GC content of blank samples compared to biological samples,
to help assess contamination levels.

```{r blank_comparison, results='asis'}
if (n_blanks == 0) {
  cat("No blank samples found in metadata.\n")
} else {
  blank_comparison <- multiqc %>%
    mutate(
      display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
      sample_group = ifelse(is.na(sample_group), "unknown", sample_group),
      type         = ifelse(sample_group == "blank", "Blank", "Biological sample")
    ) %>%
    group_by(type) %>%
    summarise(
      n            = n(),
      mean_reads   = round(mean(total_sequences, na.rm = TRUE)),
      median_reads = round(median(total_sequences, na.rm = TRUE)),
      mean_gc      = round(mean(pct_gc, na.rm = TRUE), 1),
      mean_dup     = round(mean(pct_duplicates, na.rm = TRUE), 1),
      .groups      = "drop"
    ) %>%
    mutate(
      mean_reads   = formatC(mean_reads,   format = "d", big.mark = ","),
      median_reads = formatC(median_reads, format = "d", big.mark = ",")
    )

  blank_comparison %>%
    kable(
      col.names = c("Type", "N", "Mean Reads", "Median Reads",
                    "Mean % GC", "Mean % Duplicates"),
      align = "lrrrrr"
    ) %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width        = FALSE,
      position          = "left"
    ) %>%
    print()
}
```
