---
title: "Taxonomic Profiling Report — Metagenomics Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(phyloseq)
library(vegan)
library(patchwork)
library(microbiome)

# ══════════════════════════════════════════════════════════════════════════════
# USER-ADJUSTABLE PARAMETERS
# ══════════════════════════════════════════════════════════════════════════════

# Decontamination thresholds
# Ratio filter: remove taxa where mean(blank) / mean(sample) > RATIO_THRESHOLD
RATIO_THRESHOLD <- 0.1

# Span filter: remove taxa present in >= SPAN_THRESHOLD proportion of blanks
SPAN_THRESHOLD <- 0.5

# Flag samples where unclassified reads exceed this proportion
UNCLASSIFIED_THRESHOLD <- 0.75

# Number of top taxa to display in barplots
TOP_N_TAXA <- 15
```

```{r load_config, include=FALSE}
config_lines <- readLines("../config.sh")

extract_config <- function(lines, key) {
  line <- lines[grep(paste0("^", key, "="), lines)]
  if (length(line) == 0) return(NA)
  value <- gsub(paste0(key, "=|\""), "", line[1])
  value <- gsub("#.*$", "", value)
  trimws(value)
}

cfg_type     <- extract_config(config_lines, "TYPE")
cfg_host_ref <- extract_config(config_lines, "HOST_REF")
```

```{r load_data, include=FALSE}
# ── Metadata ──────────────────────────────────────────────────────────────────
metadata_path <- "../metadata.tsv"
if (!file.exists(metadata_path)) {
  stop("metadata.tsv not found. Please copy it before rendering.")
}
metadata <- read_tsv(metadata_path, col_types = cols(.default = "c")) %>%
  rename_with(tolower)

required_cols <- c("sequence_id", "sample_name", "sample_group")
missing_cols  <- setdiff(required_cols, colnames(metadata))
if (length(missing_cols) > 0) {
  stop(paste("metadata.tsv is missing columns:", paste(missing_cols, collapse = ", ")))
}

# ── Build colour palette from sample_group ────────────────────────────────────
groups       <- unique(metadata$sample_group)
non_blank    <- groups[groups != "blank"]
base_colours <- c("#4E79A7", "#59A14F", "#F28E2B", "#B07AA1",
                  "#76B7B2", "#EDC948", "#FF9DA7", "#9C755F")
group_colours <- setNames(base_colours[seq_along(non_blank)], non_blank)
group_colours["blank"] <- "#AAAAAA"

# ── MetaPhlAn merged table ────────────────────────────────────────────────────
merged_path <- "metaphlan_merged.tsv"
if (!file.exists(merged_path)) {
  stop("metaphlan_merged.tsv not found. Has 02c_metaphlan_merge completed successfully?")
}
metaphlan_raw <- read_tsv(merged_path, skip = 1, col_types = cols(.default = "c")) %>%
  rename(clade = 1) %>%
  rename_with(~ gsub("\\.unclprofile$", "", .x), -clade)

# ── Parse MetaPhlAn table ─────────────────────────────────────────────────────
# Extract unclassified row
unclassified_row <- metaphlan_raw %>%
  filter(grepl("^UNCLASSIFIED", clade, ignore.case = TRUE))

# Keep only species-level and above rows (remove UNCLASSIFIED from main table)
metaphlan_clean <- metaphlan_raw %>%
  filter(!grepl("^UNCLASSIFIED", clade, ignore.case = TRUE))

# Convert to numeric
sample_cols <- setdiff(colnames(metaphlan_clean), "clade")
metaphlan_clean <- metaphlan_clean %>%
  mutate(across(all_of(sample_cols), as.numeric))

# ── Extract taxonomy ──────────────────────────────────────────────────────────
# MetaPhlAn clade names: k__|p__|c__|o__|f__|g__|s__
parse_taxonomy <- function(clade_string) {
  parts   <- str_split(clade_string, "\\|")[[1]]
  tax_map <- list(
    Kingdom = NA, Phylum = NA, Class = NA, Order = NA,
    Family  = NA, Genus  = NA, Species = NA
  )
  for (p in parts) {
    if (startsWith(p, "k__")) tax_map$Kingdom <- gsub("k__", "", p)
    if (startsWith(p, "p__")) tax_map$Phylum  <- gsub("p__", "", p)
    if (startsWith(p, "c__")) tax_map$Class   <- gsub("c__", "", p)
    if (startsWith(p, "o__")) tax_map$Order   <- gsub("o__", "", p)
    if (startsWith(p, "f__")) tax_map$Family  <- gsub("f__", "", p)
    if (startsWith(p, "g__")) tax_map$Genus   <- gsub("g__", "", p)
    if (startsWith(p, "s__")) tax_map$Species <- gsub("s__", "", p)
  }
  as_tibble(tax_map)
}

taxonomy <- bind_rows(lapply(metaphlan_clean$clade, parse_taxonomy))
taxonomy$clade <- metaphlan_clean$clade

# ── Build abundance matrix ────────────────────────────────────────────────────
# Keep species-level only for phyloseq
species_data <- metaphlan_clean %>%
  filter(grepl("s__", clade)) %>%
  filter(!grepl("t__", clade))  # remove strain level if present

species_tax <- taxonomy %>%
  filter(grepl("s__", clade)) %>%
  filter(!grepl("t__", clade))

# Abundance matrix: rows = species, cols = samples
abund_mat <- species_data %>%
  select(all_of(sample_cols)) %>%
  as.matrix()
rownames(abund_mat) <- species_data$clade

# Tax matrix
tax_mat <- species_tax %>%
  select(Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  as.matrix()
rownames(tax_mat) <- species_data$clade

# Sample metadata for phyloseq
sample_meta <- metadata %>%
  column_to_rownames("sequence_id")

# ── Build phyloseq object (full, including blanks) ────────────────────────────
# Align sample names between matrix and metadata
shared_samples <- intersect(colnames(abund_mat), rownames(sample_meta))
abund_mat_filt <- abund_mat[, shared_samples, drop = FALSE]
sample_meta_filt <- sample_meta[shared_samples, , drop = FALSE]

ps_full <- phyloseq(
  otu_table(abund_mat_filt, taxa_are_rows = TRUE),
  tax_table(tax_mat),
  sample_data(sample_meta_filt)
)

# ── Check for samples in metadata missing from MetaPhlAn output ───────────────
samples_missing_metaphlan <- setdiff(rownames(sample_meta), colnames(abund_mat))
if (length(samples_missing_metaphlan) > 0) {
  warning(paste(
    "The following samples are in metadata but absent from metaphlan_merged.tsv",
    "(likely failed QC or taxonomic profiling):",
    paste(samples_missing_metaphlan, collapse = ", ")
  ))
}

# Summary counts — based on what is actually in the MetaPhlAn output
n_total   <- nsamples(ps_full)
n_samples <- sum(sample_data(ps_full)$sample_group != "blank", na.rm = TRUE)
n_blanks  <- sum(sample_data(ps_full)$sample_group == "blank", na.rm = TRUE)
n_species <- nrow(species_data)
```

---

# 1. Project Overview

```{r project_summary}
tibble(
  Parameter = c(
    "Sequencing type", "Host reference",
    "Total samples", "Biological samples", "Blank samples",
    "Species detected (pre-decontamination)",
    "Decontamination ratio threshold",
    "Decontamination span threshold",
    "Report generated"
  ),
  Value = c(
    toupper(cfg_type), cfg_host_ref,
    n_total, n_samples, n_blanks,
    n_species,
    RATIO_THRESHOLD,
    SPAN_THRESHOLD,
    as.character(Sys.Date())
  )
) %>%
  kable(col.names = c("Parameter", "Value"), align = "ll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")
```

---

# 2. Unclassified Reads

Proportion of reads that could not be classified by MetaPhlAn4 per sample.
Samples exceeding `r scales::percent(UNCLASSIFIED_THRESHOLD)` unclassified reads are flagged.

```{r unclassified_plot}
if (nrow(unclassified_row) > 0) {
  uncl_data <- unclassified_row %>%
    select(all_of(sample_cols)) %>%
    pivot_longer(everything(), names_to = "sequence_id", values_to = "pct_unclassified") %>%
    mutate(pct_unclassified = as.numeric(pct_unclassified) / 100) %>%
    left_join(metadata, by = "sequence_id") %>%
    mutate(
      display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
      sample_group = ifelse(is.na(sample_group), "unknown", sample_group),
      flagged      = pct_unclassified > UNCLASSIFIED_THRESHOLD
    )

  plot_colours <- c(group_colours, "flagged" = "#E15759")

  ggplot(uncl_data, aes(
      x    = reorder(display_name, pct_unclassified),
      y    = pct_unclassified,
      fill = ifelse(flagged, "flagged", sample_group)
    )) +
    geom_col() +
    geom_hline(yintercept = UNCLASSIFIED_THRESHOLD,
               linetype = "dashed", colour = "grey40") +
    scale_fill_manual(values = plot_colours, name = NULL) +
    scale_y_continuous(labels = scales::percent) +
    coord_flip() +
    labs(x = NULL, y = "% Unclassified", title = "Unclassified reads per sample") +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom")
} else {
  cat("No UNCLASSIFIED row found in merged MetaPhlAn table.\n")
}
```

```{r unclassified_flags, results='asis'}
if (nrow(unclassified_row) > 0) {
  flagged <- uncl_data %>%
    filter(flagged) %>%
    select(display_name, sample_group, pct_unclassified) %>%
    mutate(pct_unclassified = scales::percent(pct_unclassified, accuracy = 0.1))

  if (nrow(flagged) > 0) {
    cat(paste0("**⚠ ", nrow(flagged),
               " sample(s) exceed the unclassified threshold (",
               scales::percent(UNCLASSIFIED_THRESHOLD), "):**\n\n"))
    flagged %>%
      kable(col.names = c("Sample", "Group", "% Unclassified"), align = "llr") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                    full_width = FALSE, position = "left") %>%
      row_spec(seq_len(nrow(flagged)), background = "#f8d7da") %>%
      print()
  } else {
    cat(paste0("✓ All samples are below the unclassified threshold (",
               scales::percent(UNCLASSIFIED_THRESHOLD), ").\n"))
  }
}
```

---

# 3. Decontamination

Contaminants are identified using the MicrobIEM ratio and span filters applied at
species level using blank samples as negative controls.

- **Ratio filter**: a taxon is flagged if its mean relative abundance in blanks divided
  by its mean relative abundance in biological samples exceeds `r RATIO_THRESHOLD`.
- **Span filter**: a taxon is flagged if it is present in ≥ `r scales::percent(SPAN_THRESHOLD)`
  of blank samples.

Taxa flagged by either filter are removed. The cleaned phyloseq object is saved to disk.

```{r decontamination}
if (n_blanks == 0) {
  cat("No blank samples found in metadata. Decontamination step skipped.\n")
  ps_clean <- subset_samples(ps_full, sample_group != "blank")
} else {
  # Split phyloseq into blanks and samples
  ps_blanks  <- subset_samples(ps_full, sample_group == "blank")
  ps_samples <- subset_samples(ps_full, sample_group != "blank")

  abund_blank   <- as.data.frame(otu_table(ps_blanks))
  abund_samples <- as.data.frame(otu_table(ps_samples))

  # Convert to relative abundance if not already
  rel_blank   <- sweep(abund_blank,   2, colSums(abund_blank),   "/")
  rel_samples <- sweep(abund_samples, 2, colSums(abund_samples), "/")

  # Replace NaN from zero-sum columns
  rel_blank[is.nan(as.matrix(rel_blank))]     <- 0
  rel_samples[is.nan(as.matrix(rel_samples))] <- 0

  mean_blank   <- rowMeans(rel_blank,   na.rm = TRUE)
  mean_samples <- rowMeans(rel_samples, na.rm = TRUE)

  # Ratio filter
  ratio <- ifelse(mean_samples > 0, mean_blank / mean_samples, Inf)
  ratio_flag <- ratio > RATIO_THRESHOLD

  # Span filter: proportion of blanks where taxon is present
  span <- rowSums(abund_blank > 0) / ncol(abund_blank)
  span_flag <- span >= SPAN_THRESHOLD

  # Combined flag
  contaminant_flag <- ratio_flag | span_flag

  n_contaminants   <- sum(contaminant_flag)
  contaminant_taxa <- names(contaminant_flag[contaminant_flag])

  # Remove contaminants and blanks from phyloseq
  ps_clean <- subset_samples(ps_full, sample_group != "blank")
  if (n_contaminants > 0) {
    taxa_to_keep <- taxa_names(ps_clean)[!taxa_names(ps_clean) %in% contaminant_taxa]
    ps_clean     <- prune_taxa(taxa_to_keep, ps_clean)
  }

  # Remove taxa with zero abundance after blank removal
  ps_clean <- prune_taxa(taxa_sums(ps_clean) > 0, ps_clean)
}

n_taxa_clean   <- ntaxa(ps_clean)
n_taxa_removed <- n_species - n_taxa_clean

# ── Sample filtering ──────────────────────────────────────────────────────────
# 1. Remove samples exceeding the unclassified threshold
samples_high_uncl <- character(0)
if (nrow(unclassified_row) > 0) {
  samples_high_uncl <- uncl_data %>%
    filter(flagged) %>%
    pull(sequence_id)
  if (length(samples_high_uncl) > 0) {
    ps_clean <- subset_samples(
      ps_clean,
      !rownames(sample_data(ps_clean)) %in% samples_high_uncl
    )
  }
}

# 2. Remove samples where all taxa were removed by decontamination
sample_sums_clean  <- sample_sums(ps_clean)
samples_empty      <- names(sample_sums_clean[sample_sums_clean == 0])
if (length(samples_empty) > 0) {
  ps_clean <- subset_samples(ps_clean, sample_sums(ps_clean) > 0)
}

n_samples_removed_uncl  <- length(samples_high_uncl)
n_samples_removed_empty <- length(samples_empty)
n_samples_final         <- nsamples(ps_clean)

# ── Rescale to compositional (TSS) ───────────────────────────────────────────
ps_clean <- transform_sample_counts(ps_clean, function(x) x / sum(x) * 100)

# Save final cleaned and rescaled phyloseq object
saveRDS(ps_clean, file = "ps_clean.rds")
```

```{r decontamination_summary, results='asis'}
if (n_blanks > 0) {
  tibble(
    Metric = c(
      "Species before decontamination",
      "Contaminants removed (ratio filter)",
      "Contaminants removed (span filter)",
      "Total taxa removed",
      "Species after decontamination"
    ),
    Value = c(
      n_species,
      sum(ratio_flag),
      sum(span_flag),
      n_taxa_removed,
      n_taxa_clean
    )
  ) %>%
    kable(col.names = c("Metric", "Value"), align = "lr") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    print()

  if (n_contaminants > 0) {
    cat("\n**Removed taxa:**\n\n")
    tibble(
      Taxon  = contaminant_taxa,
      Ratio  = round(ratio[contaminant_taxa], 4),
      Span   = round(span[contaminant_taxa], 2),
      Reason = case_when(
        ratio_flag[contaminant_taxa] & span_flag[contaminant_taxa] ~ "ratio + span",
        ratio_flag[contaminant_taxa] ~ "ratio",
        TRUE ~ "span"
      )
    ) %>%
      arrange(desc(Ratio)) %>%
      kable(col.names = c("Taxon", "Ratio (blank/sample)",
                          "Span (prop. blanks)", "Filter"),
            align = "lrrl") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                    full_width = TRUE) %>%
      row_spec(seq_len(n_contaminants), background = "#fff3cd") %>%
      print()
  } else {
    cat("✓ No contaminants identified with the current thresholds.\n")
  }
}
```

## 3.4 Sample Filtering and Compositional Rescaling

Samples are excluded from downstream analysis if they exceed the unclassified
threshold (`r scales::percent(UNCLASSIFIED_THRESHOLD)`) or if all their taxa were
removed during decontamination. Remaining samples are then rescaled to
compositional relative abundances (TSS, summing to 100%) before downstream analyses.

```{r sample_filtering_summary, results='asis'}
tibble(
  Metric = c(
    "Samples before filtering",
    "Removed (unclassified > threshold)",
    "Removed (all taxa depleted by decontamination)",
    "Samples retained for downstream analysis",
    "Normalisation"
  ),
  Value = c(
    n_samples,
    n_samples_removed_uncl,
    n_samples_removed_empty,
    n_samples_final,
    "TSS (relative abundance, sum to 100%)"
  )
) %>%
  kable(col.names = c("Metric", "Value"), align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left") %>%
  print()

if (n_samples_removed_uncl > 0) {
  cat("\n**⚠ Samples removed (high unclassified proportion):**\n\n")
  tibble(sequence_id = samples_high_uncl) %>%
    left_join(metadata, by = "sequence_id") %>%
    select(sequence_id, sample_name, sample_group) %>%
    kable(col.names = c("Sequence ID", "Sample Name", "Group"), align = "lll") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    row_spec(seq_len(n_samples_removed_uncl), background = "#f8d7da") %>%
    print()
}

if (n_samples_removed_empty > 0) {
  cat("\n**⚠ Samples removed (all taxa depleted after decontamination):**\n\n")
  tibble(sequence_id = samples_empty) %>%
    left_join(metadata, by = "sequence_id") %>%
    select(sequence_id, sample_name, sample_group) %>%
    kable(col.names = c("Sequence ID", "Sample Name", "Group"), align = "lll") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    row_spec(seq_len(n_samples_removed_empty), background = "#f8d7da") %>%
    print()
}

if (n_samples_removed_uncl == 0 && n_samples_removed_empty == 0) {
  cat("✓ No samples removed. All samples retained for downstream analysis.\n")
}

cat(paste0("\nFinal cleaned and rescaled phyloseq object saved to: ",
           "`02_taxonomic_profiling/ps_clean.rds`\n"))
```

---

# 4. Alpha Diversity

Observed species richness, Shannon entropy, and Pielou's evenness per sample,
computed on the decontaminated species-level profiles.

```{r alpha_diversity}
# Compute alpha diversity metrics
alpha_div <- microbiome::alpha(ps_clean, index = c("Observed", "Shannon", "Pielou")) %>%
  rownames_to_column("sequence_id") %>%
  left_join(metadata, by = "sequence_id") %>%
  mutate(
    display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group)
  ) %>%
  rename("Observed"="observed", "Shannon"="diversity_shannon", "Pielou"="evenness_pielou")

# Reshape for faceted plot
alpha_long <- alpha_div %>%
  select(display_name, sample_group, Observed, Shannon, Pielou) %>%
  pivot_longer(
    cols      = c(Observed, Shannon, Pielou),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  mutate(metric = factor(metric, levels = c("Observed", "Shannon", "Pielou")))

metric_labels <- c(
  "Observed" = "Observed Richness",
  "Shannon"  = "Shannon Entropy",
  "Pielou"   = "Pielou Evenness"
)

ggplot(alpha_long, aes(x = sample_group, y = value, fill = sample_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8, aes(colour = sample_group)) +
  scale_fill_manual(values  = group_colours, name = NULL) +
  scale_colour_manual(values = group_colours, name = NULL) +
  facet_wrap(~metric, scales = "free_y", labeller = labeller(metric = metric_labels)) +
  labs(x = NULL, y = NULL, title = "Alpha Diversity by Sample Group") +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 30, hjust = 1),
    strip.background = element_rect(fill = "#f0f0f0")
  )
```

```{r alpha_table}
alpha_div %>%
  select(display_name, sample_group, Observed, Shannon, Pielou) %>%
  mutate(
    Shannon = round(Shannon, 3),
    Pielou  = round(Pielou,  3)
  ) %>%
  arrange(sample_group, display_name) %>%
  kable(
    col.names = c("Sample", "Group", "Observed Richness",
                  "Shannon", "Pielou Evenness"),
    align = "llrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE)
```

---

# 5. Beta Diversity

Principal Coordinates Analysis (PCoA) based on Bray-Curtis dissimilarity computed
on species-level relative abundances after decontamination.

```{r beta_diversity}
# Compute Bray-Curtis distance
abund_clean <- as.data.frame(t(otu_table(ps_clean)))  # samples as rows
bc_dist     <- vegdist(abund_clean, method = "bray")

# PCoA
pcoa_res  <- cmdscale(bc_dist, eig = TRUE, k = 2)
pct_var   <- round(pcoa_res$eig / sum(pcoa_res$eig[pcoa_res$eig > 0]) * 100, 1)

pcoa_df <- as.data.frame(pcoa_res$points) %>%
  setNames(c("PC1", "PC2")) %>%
  rownames_to_column("sequence_id") %>%
  left_join(metadata, by = "sequence_id") %>%
  mutate(
    display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group)
  )

ggplot(pcoa_df, aes(x = PC1, y = PC2,
                    colour = sample_group, label = display_name)) +
  geom_point(size = 3, alpha = 0.85) +
  stat_ellipse(aes(group = sample_group), type = "t", level = 0.95,
               linetype = "dashed", linewidth = 0.5) +
  scale_colour_manual(values = group_colours, name = "Group") +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 10, show.legend = FALSE) +
  labs(
    x = paste0("PC1 (", pct_var[1], "%)"),
    y = paste0("PC2 (", pct_var[2], "%)"),
    title = "PCoA — Bray-Curtis Dissimilarity (species level)"
  ) +
  theme_bw(base_size = 11)
```

```{r permanova, results='asis'}
# PERMANOVA test
meta_clean <- as(sample_data(ps_clean), "data.frame")
if (length(unique(meta_clean$sample_group)) > 1) {
  perm <- adonis2(bc_dist ~ sample_group, data = meta_clean, permutations = 999)
  cat("**PERMANOVA result (Bray-Curtis, 999 permutations):**\n\n")
  perm %>%
    as.data.frame() %>%
    rownames_to_column("Term") %>%
    mutate(across(where(is.numeric), ~round(.x, 4))) %>%
    kable(align = "lrrrrr") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left") %>%
    print()
} else {
  cat("Only one sample group detected — PERMANOVA not computed.\n")
}
```

---

# 6. Taxonomic Composition

Top `r TOP_N_TAXA` taxa by mean relative abundance across biological samples,
after decontamination. Blanks are excluded.

```{r tax_barplot_function, include=FALSE}
make_tax_barplot <- function(ps_obj, level, top_n, group_cols) {
  # Aggregate at requested taxonomic level
  ps_level <- tax_glom(ps_obj, taxrank = level, NArm = FALSE)

  # Convert to relative abundance
  ps_rel <- transform_sample_counts(ps_level, function(x) x / sum(x) * 100)

  # Get top N taxa by mean abundance
  mean_abund  <- rowMeans(otu_table(ps_rel))
  top_taxa    <- names(sort(mean_abund, decreasing = TRUE))[seq_len(min(top_n, length(mean_abund)))]

  # Melt to long format
  abund_long <- as.data.frame(t(otu_table(ps_rel))) %>%
    rownames_to_column("sequence_id") %>%
    pivot_longer(-sequence_id, names_to = "taxon_id", values_to = "abundance") %>%
    left_join(
      as.data.frame(tax_table(ps_rel)) %>% rownames_to_column("taxon_id"),
      by = "taxon_id"
    ) %>%
    left_join(
      as(sample_data(ps_rel), "data.frame") %>% rownames_to_column("sequence_id"),
      by = "sequence_id"
    ) %>%
    mutate(
      taxon_label  = ifelse(is.na(.data[[level]]) | .data[[level]] == "",
                            "Unassigned", .data[[level]]),
      display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
      in_top       = taxon_id %in% top_taxa,
      taxon_plot   = ifelse(in_top, taxon_label, "Other")
    )

  # Get ordered list of top taxa for colour assignment
  top_labels <- abund_long %>%
    filter(in_top) %>%
    group_by(taxon_plot) %>%
    summarise(mean_abund = mean(abundance), .groups = "drop") %>%
    arrange(desc(mean_abund)) %>%
    pull(taxon_plot)

  all_labels   <- c(top_labels, "Other")
  n_top        <- length(top_labels)
  tax_colours  <- setNames(
    c(colorRampPalette(c("#4E79A7","#F28E2B","#59A14F","#E15759",
                         "#76B7B2","#EDC948","#B07AA1","#FF9DA7",
                         "#9C755F","#BAB0AC"))(n_top), "#DDDDDD"),
    all_labels
  )

  # Summarise by sample
  plot_data <- abund_long %>%
    group_by(display_name, sample_group, taxon_plot) %>%
    summarise(abundance = sum(abundance), .groups = "drop") %>%
    mutate(taxon_plot = factor(taxon_plot, levels = rev(all_labels)))

  ggplot(plot_data, aes(x = display_name, y = abundance, fill = taxon_plot)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = tax_colours, name = level) +
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    labs(x = NULL, y = "Relative Abundance", title = paste0("Top ", top_n, " ", level)) +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x    = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      legend.text     = element_text(size = 8)
    ) +
    guides(fill = guide_legend(reverse = TRUE, ncol = 1))
}
```

## 6.1 Family Level

```{r barplot_family, fig.height=6}
make_tax_barplot(ps_clean, level = "Family", top_n = TOP_N_TAXA,
                 group_cols = group_colours)
```

## 6.2 Genus Level

```{r barplot_genus, fig.height=6}
make_tax_barplot(ps_clean, level = "Genus", top_n = TOP_N_TAXA,
                 group_cols = group_colours)
```

## 6.3 Species Level

```{r barplot_species, fig.height=6}
make_tax_barplot(ps_clean, level = "Species", top_n = TOP_N_TAXA,
                 group_cols = group_colours)
```
