---
title: "Fungal Analysis Report — Metagenomics Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 6
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)

# ══════════════════════════════════════════════════════════════════════════════
# USER-ADJUSTABLE PARAMETERS
# ══════════════════════════════════════════════════════════════════════════════

# Minimum number of reads assigned to a taxon in a sample to be considered
# present (avoids Kraken2 false positives from background noise)
MIN_FUNGAL_READS <- 10

# Number of top genera shown in the relative abundance barplot
TOP_N_TAXA <- 15
```

```{r load_config, include=FALSE}
config_lines <- readLines("../config.sh")

extract_config <- function(lines, key) {
  line <- lines[grep(paste0("^", key, "="), lines)]
  if (length(line) == 0) return(NA)
  value <- gsub(paste0(key, "=|\""), "", line[1])
  value <- gsub("#.*$", "", value)
  trimws(value)
}

cfg_type    <- extract_config(config_lines, "TYPE")
cfg_out_dir <- extract_config(config_lines, "OUT_DIR")
kraken_dir  <- file.path(cfg_out_dir, "02_taxonomic_profiling/kraken2")
```

```{r load_metadata, include=FALSE}
# ── Metadata ──────────────────────────────────────────────────────────────────
metadata_path <- "../metadata.tsv"
if (!file.exists(metadata_path)) {
  stop("metadata.tsv not found. Please add it to 04_aggregated/ before rendering.")
}
metadata <- read_tsv(metadata_path, col_types = cols(.default = "c")) %>%
  rename_with(tolower)

required_cols <- c("sequence_id", "sample_name", "sample_group")
missing_cols  <- setdiff(required_cols, colnames(metadata))
if (length(missing_cols) > 0) {
  stop(paste("metadata.tsv is missing columns:", paste(missing_cols, collapse = ", ")))
}

# ── Build colour palette from sample_group ────────────────────────────────────
groups        <- unique(metadata$sample_group)
non_blank     <- groups[groups != "blank"]
base_colours  <- c("#4E79A7", "#59A14F", "#F28E2B", "#B07AA1",
                   "#76B7B2", "#EDC948", "#FF9DA7", "#9C755F")
group_colours <- setNames(base_colours[seq_along(non_blank)], non_blank)
group_colours["blank"] <- "#AAAAAA"
```

```{r parse_kraken, include=FALSE}
# ── Parse Kraken2 report files ────────────────────────────────────────────────
# Kraken2 report columns:
#   1: % reads covered by clade
#   2: reads covered by clade (rooted)
#   3: reads assigned directly to taxon
#   4: rank code (U, R, D, K, P, C, O, F, G, S...)
#   5: NCBI taxon ID
#   6: name (indented with spaces to show hierarchy)

parse_kraken_report <- function(filepath) {
  lines <- readLines(filepath)
  df <- read_tsv(filepath,
                 col_names = c("pct", "reads_rooted", "reads_direct",
                               "rank", "taxid", "name"),
                 col_types = "dddccc")
  df$indent <- nchar(df$name) - nchar(trimws(df$name, "left"))
  df$name   <- trimws(df$name)
  df
}

# Find the Fungi kingdom block in a parsed report
extract_fungi <- function(df) {
  fungi_row <- which(df$name == "Fungi" & df$rank == "K")
  if (length(fungi_row) == 0) return(NULL)

  fungi_indent <- df$indent[fungi_row]

  # Collect all rows after the Fungi row with deeper indentation
  end_row <- nrow(df)
  for (i in (fungi_row + 1):nrow(df)) {
    if (df$indent[i] <= fungi_indent) {
      end_row <- i - 1
      break
    }
  }
  df[fungi_row:end_row, ]
}

# ── Read all Kraken2 reports ──────────────────────────────────────────────────
kraken_files <- list.files(kraken_dir, pattern = "_kraken_report\\.txt$",
                           full.names = TRUE)

if (length(kraken_files) == 0) {
  stop(paste("No Kraken2 report files found in", kraken_dir,
             "-- has 02a_taxonomic_profiling completed?"))
}

# Extract sample IDs from filenames
sample_ids <- gsub("_kraken_report\\.txt$", "", basename(kraken_files))

# Parse all reports and extract fungal blocks
fungal_list <- lapply(seq_along(kraken_files), function(i) {
  df  <- parse_kraken_report(kraken_files[i])
  fun <- extract_fungi(df)
  if (is.null(fun)) return(NULL)
  fun$sequence_id <- sample_ids[i]

  # Also capture total classified reads for relative abundance calculation
  total_row <- df %>% filter(rank == "R")
  fun$total_classified <- if (nrow(total_row) > 0) total_row$reads_rooted[1] else NA
  fun
})

# Total reads per sample (from root row) for fungal proportion calculation
total_reads_list <- lapply(seq_along(kraken_files), function(i) {
  df <- parse_kraken_report(kraken_files[i])
  root_row <- df %>% filter(rank == "R")
  fungi_row <- df %>% filter(name == "Fungi", rank == "K")
  tibble(
    sequence_id     = sample_ids[i],
    total_classified = if (nrow(root_row)  > 0) root_row$reads_rooted[1]  else 0,
    fungal_reads    = if (nrow(fungi_row) > 0) fungi_row$reads_rooted[1] else 0
  )
})
total_reads_df <- bind_rows(total_reads_list) %>%
  left_join(metadata, by = "sequence_id") %>%
  mutate(
    display_name    = ifelse(is.na(sample_name), sequence_id, sample_name),
    sample_group    = ifelse(is.na(sample_group), "unknown", sample_group),
    pct_fungal      = ifelse(total_classified > 0,
                             fungal_reads / total_classified * 100, 0),
    has_fungi       = fungal_reads >= MIN_FUNGAL_READS
  )

# Combine all fungal blocks
fungal_all <- bind_rows(Filter(Negate(is.null), fungal_list))

# ── Extract genus-level data ──────────────────────────────────────────────────
fungal_genus <- fungal_all %>%
  filter(rank == "G") %>%
  filter(reads_rooted >= MIN_FUNGAL_READS) %>%
  select(sequence_id, genus = name, reads_rooted, total_classified) %>%
  left_join(metadata, by = "sequence_id") %>%
  mutate(
    display_name = ifelse(is.na(sample_name), sequence_id, sample_name),
    sample_group = ifelse(is.na(sample_group), "unknown", sample_group),
    rel_abund    = ifelse(!is.na(total_classified) & total_classified > 0,
                          reads_rooted / total_classified * 100, 0)
  )

# Summary counts
n_total        <- nrow(metadata)
n_with_fungi   <- sum(total_reads_df$has_fungi, na.rm = TRUE)
n_genera       <- length(unique(fungal_genus$genus))
n_kraken_files <- length(kraken_files)
```

---

# 1. Project Overview

```{r project_summary}
tibble(
  Parameter = c(
    "Sequencing type",
    "Kraken2 reports found",
    "Total samples in metadata",
    "Samples with fungal signal",
    "Samples without fungal signal",
    "Fungal genera detected (across all samples)",
    "Minimum reads threshold",
    "Report generated"
  ),
  Value = c(
    toupper(cfg_type),
    n_kraken_files,
    n_total,
    n_with_fungi,
    n_total - n_with_fungi,
    n_genera,
    MIN_FUNGAL_READS,
    as.character(Sys.Date())
  )
) %>%
  kable(col.names = c("Parameter", "Value"), align = "ll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")
```

---

# 2. Fungal Read Proportion

Proportion of total classified reads attributed to the Fungi kingdom per sample.
Samples with fewer than `r MIN_FUNGAL_READS` fungal reads are considered to have
no detectable fungal signal.

```{r fungal_proportion_plot, fig.height=6}
ggplot(total_reads_df,
       aes(x    = reorder(display_name, pct_fungal),
           y    = pct_fungal,
           fill = sample_group)) +
  geom_col() +
  geom_hline(yintercept = 0, colour = "grey40") +
  scale_fill_manual(values = group_colours, name = NULL) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "% Classified Reads (Fungi)",
    title = "Proportion of classified reads attributed to Fungi per sample"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")
```

```{r fungal_proportion_table}
total_reads_df %>%
  select(display_name, sample_group, total_classified, fungal_reads, pct_fungal) %>%
  mutate(
    total_classified = formatC(total_classified, format = "d", big.mark = ","),
    fungal_reads     = formatC(fungal_reads,     format = "d", big.mark = ","),
    pct_fungal       = round(pct_fungal, 3)
  ) %>%
  arrange(sample_group, display_name) %>%
  kable(
    col.names = c("Sample", "Group", "Total Classified Reads",
                  "Fungal Reads", "% Fungal"),
    align = "llrrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(which(total_reads_df$has_fungi == FALSE), background = "#f0f0f0")
```

---

# 3. Fungal Genus Presence / Absence

Presence/absence heatmap at genus level across all samples. A taxon is considered
present if it has at least `r MIN_FUNGAL_READS` reads in that sample. All samples
are shown including those with no detected fungi.

```{r presence_absence_heatmap, fig.height=8}
# Build full presence/absence matrix: all samples x all detected genera
all_genera   <- unique(fungal_genus$genus)
all_samples  <- total_reads_df %>%
  select(sequence_id, display_name, sample_group) %>%
  arrange(sample_group, display_name)

pa_matrix <- expand_grid(
    sequence_id = all_samples$sequence_id,
    genus       = all_genera
  ) %>%
  left_join(
    fungal_genus %>% select(sequence_id, genus, reads_rooted),
    by = c("sequence_id", "genus")
  ) %>%
  mutate(present = !is.na(reads_rooted) & reads_rooted >= MIN_FUNGAL_READS) %>%
  left_join(all_samples, by = "sequence_id")

# Order genera by prevalence (most prevalent first)
genus_order <- pa_matrix %>%
  group_by(genus) %>%
  summarise(prevalence = sum(present), .groups = "drop") %>%
  arrange(desc(prevalence)) %>%
  pull(genus)

pa_matrix <- pa_matrix %>%
  mutate(genus = factor(genus, levels = genus_order))

ggplot(pa_matrix, aes(x = genus, y = display_name, fill = present)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = c("TRUE" = "#4E79A7", "FALSE" = "#F0F0F0"),
    labels = c("TRUE" = "Present", "FALSE" = "Absent"),
    name   = NULL
  ) +
  facet_grid(sample_group ~ ., scales = "free_y", space = "free_y") +
  labs(
    x     = "Genus",
    y     = NULL,
    title = paste0("Fungal genus presence/absence (min. ",
                   MIN_FUNGAL_READS, " reads)")
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, face = "italic"),
    legend.position  = "bottom",
    strip.background = element_rect(fill = "#f0f0f0"),
    strip.text.y     = element_text(angle = 0)
  )
```

```{r prevalence_table}
pa_matrix %>%
  group_by(genus) %>%
  summarise(
    n_present    = sum(present),
    pct_present  = round(n_present / n_distinct(sequence_id) * 100, 1),
    .groups      = "drop"
  ) %>%
  arrange(desc(n_present)) %>%
  kable(
    col.names = c("Genus", "# Samples Present", "% Samples Present"),
    align     = "lrr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "left")
```

---

# 4. Fungal Relative Abundance

Top `r TOP_N_TAXA` fungal genera by mean relative abundance across samples with
detectable fungal signal. Relative abundance is expressed as a proportion of total
classified reads. Samples with no detected fungi are excluded.

```{r fungal_barplot, fig.height=7}
# Samples with fungal signal
samples_with_fungi <- total_reads_df %>%
  filter(has_fungi) %>%
  pull(sequence_id)

if (length(samples_with_fungi) == 0) {
  cat("No samples with detectable fungal signal above the minimum read threshold.\n")
} else {
  # Get top N genera by mean relative abundance across samples with fungi
  top_genera <- fungal_genus %>%
    filter(sequence_id %in% samples_with_fungi) %>%
    group_by(genus) %>%
    summarise(mean_rel = mean(rel_abund), .groups = "drop") %>%
    arrange(desc(mean_rel)) %>%
    slice_head(n = TOP_N_TAXA) %>%
    pull(genus)

  # Build plot data — all genera not in top N collapsed to "Other"
  plot_data <- fungal_genus %>%
    filter(sequence_id %in% samples_with_fungi) %>%
    mutate(genus_plot = ifelse(genus %in% top_genera, genus, "Other")) %>%
    group_by(sequence_id, display_name, sample_group, genus_plot) %>%
    summarise(rel_abund = sum(rel_abund), .groups = "drop") %>%
    mutate(
      genus_plot = factor(genus_plot,
                          levels = c(rev(top_genera), "Other"))
    )

  # Colour palette for taxa
  n_top      <- length(top_genera)
  tax_colours <- setNames(
    c(colorRampPalette(c("#4E79A7", "#F28E2B", "#59A14F", "#E15759",
                         "#76B7B2", "#EDC948", "#B07AA1", "#FF9DA7",
                         "#9C755F", "#BAB0AC"))(n_top), "#DDDDDD"),
    c(rev(top_genera), "Other")
  )

  ggplot(plot_data,
         aes(x    = display_name,
             y    = rel_abund,
             fill = genus_plot)) +
    geom_col(position = "stack") +
    scale_fill_manual(values = tax_colours, name = "Genus") +
    scale_y_continuous(labels = function(x) paste0(x, "%")) +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free_x") +
    labs(
      x     = NULL,
      y     = "% Classified Reads",
      title = paste0("Top ", TOP_N_TAXA,
                     " fungal genera by relative abundance")
    ) +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x      = element_text(angle = 45, hjust = 1),
      legend.position  = "right",
      legend.text      = element_text(size = 8, face = "italic"),
      strip.background = element_rect(fill = "#f0f0f0")
    ) +
    guides(fill = guide_legend(reverse = TRUE, ncol = 1))
}
```

```{r fungal_abundance_table}
if (length(samples_with_fungi) > 0) {
  fungal_genus %>%
    filter(sequence_id %in% samples_with_fungi,
           genus %in% top_genera) %>%
    select(display_name, sample_group, genus, reads_rooted, rel_abund) %>%
    mutate(
      reads_rooted = formatC(reads_rooted, format = "d", big.mark = ","),
      rel_abund    = round(rel_abund, 4)
    ) %>%
    arrange(sample_group, display_name, desc(rel_abund)) %>%
    kable(
      col.names = c("Sample", "Group", "Genus",
                    "Reads (rooted)", "% Classified Reads"),
      align = "llrr"
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE)
}
```

---

# 5. Notes on Interpretation

```{r interpretation_notes, results='asis'}
cat(paste0(
  "- Fungal relative abundances are expressed as a proportion of **total classified reads**, ",
  "not total reads. Samples with a low overall classification rate may therefore appear to ",
  "have more fungi than they truly contain.\n",
  "- A minimum threshold of **", MIN_FUNGAL_READS, " reads** per taxon per sample is applied ",
  "to reduce false positives from Kraken2 background noise. This threshold can be adjusted ",
  "in the `USER-ADJUSTABLE PARAMETERS` section at the top of this report.\n",
  "- Common laboratory contaminant fungi (e.g. *Malassezia*, *Candida*, *Aspergillus*) ",
  "may appear in negative controls. Cross-reference the presence/absence heatmap with ",
  "blank samples before interpreting results.\n",
  "- Kraken2 without Bracken re-estimation is used here intentionally. Bracken is not ",
  "applied as it can distort abundances when a large proportion of reads remain unclassified ",
  "or are classified only at high taxonomic levels, which is common in food metagenomes.\n",
  "- Species-level resolution is available in the Kraken2 reports but is not shown here ",
  "due to lower reliability at that level without abundance re-estimation. Contact the ",
  "Bioinformatics Core if species-level analysis is required."
))
```
